Automatically Splitting a Two-Stage Lambda Calculus

### 2015-07-10 paper/relatedwork.tex
background

\paragraph{Precomputation techniques.}
Perhaps one of the most studied examples of precomputation is
partial evaluation.  Partial evaluation distinguishes between static
(compile-time) and dynamic (runtime) stages. Given a program and values 
for all static inputs, partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene, who in 1952 proved that a multivariate program (function)
$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
%Kleene did not intend to improve efficiency, however.  Lombardi
%appears to be the first person to use the term in his work on
%incremental computation~\cite{Lombardi67}.
There has since been an enormous body of work on partial evaluation,
and we refer the reader to the book by Jones, Gommard, and Sestoft for
a comprehensive discussion of work until the early 90's~\cite{JGS93}.

@@

@@

\TODO move this

In the partial evaluation community, metastatic staging transforms (called \emph{program bifurcation} and \emph{binding time separation}) 
similar to splitting have been investigated 
as a preprocessing step necessary for simplifying the specialization phase of self-applied partial evaluators. 
For instance, Mogensen\,\cite{Mogensen89a} and Niel et al.\,\cite{DeNiel91} defined transformations for a set of first-order recursive functions.

Those works operate on the results of binding time analysis

and so much of their binding-time specification just amounts to a type system.
Formulations like \lang\ (namely the two-level system of~\cite{NN92-twolevel}) were known at the time,
but~\cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.

Davies essentially solved this limitation in~\cite{davies96}

Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b}.

@@

Algorithms for stage splitting have appeared in the literature under the names 
{\em pass separation}\,\cite{JS86-staging} and {\em data specialization}\,\cite{knoblock96}.

Splitting algorithms have also been a topic of interest in programming
systems for computer graphics, where to achieve high performance,
programs are typically manually separated into components by frequency
of execution. (Indeed, separation into a pre-determined set of
per-frequency modules: e.g., per frame, per pixel, per triangle, is
\emph{required} by the execution model of modern graphics
architectures.)  The software engineering challenges of modifying
graphics programs written in a split form have spurred interest in
deferring the task of splitting to a compiler\,\cite{Foley:2011}.
While early work sought to determine the staging of programs
automatically via binding time analysis\,\cite{knoblock96}, more
recent designs, like \lang, require explicit staging by the
programmer\,\cite{Proudfoot:2001,Foley:2011,He:2014}.  However,
splitting algorithms developed in all of these efforts have been
limited to simple, imperative graphics shading languages that lack
even simple features such as functions.

>>>>

\paragraph{Precomputation techniques.}
Perhaps one of the most studied examples of precomputation is
partial evaluation.  Partial evaluation distinguishes between static
(compile-time) and dynamic (runtime) stages. Given a program and values 
for all static inputs, partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52}.
%Kleene did not intend to improve efficiency, however.  Lombardi
%appears to be the first person to use the term in his work on
%incremental computation~\cite{Lombardi67}.
There has since been an enormous body of work on partial evaluation,
and we refer the reader to the book by Jones, Gommard, and Sestoft for
a comprehensive discussion of work until the early 90's~\cite{JGS93}.

@@

\emph{Program bifurcation}\,\cite{DeNiel91} (also called \emph{binding time separation}\,\cite{Mogensen89a})
is a metastatic staging transformations
used as a preprocessing step to simplify the specialization phase of self-applied partial evaluators. 
Bifurcation is similar to stage splitting in that it factors the input program
into two versions based on the stages of the input.
However, in bifurcation there is no building of data structures to be use in the resumer;
instead, the second stage version of the code also has 
access to the first stage inputs and just recomputes whatever it needs.
That difference aside, the grammar-based binding-time specifications used in bifurcation
were capable of descring data structures with purely-static, purely-dynamic, and mixed stage content,
much like the type system of \lang.

@@

@@

Algorithms for stage splitting have appeared in the literature under the names 
{\em pass separation}\,\cite{JS86-staging} and {\em data specialization}\,\cite{knoblock96}.
Perhaps the closest work to ours is the algorithm for data
specialization given by Knoblock and Ruf~\cite{knoblock96}, which
also seeks to statically split an explicitly staged program into two
stages.  However, they only consider a simple first-order language and
straight-line programs. Their work also treats all computations
guarded by second-stage conditionals as second-stage computations,
which would prevent optimization (via splitting) of programs such as
quickselect.

Splitting algorithms have also been a topic of interest in
programming systems for computer graphics, where, to achieve
high performance, programs are typically manually separated into
components by frequency of execution (Indeed, separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges of modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly-staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass separation to the compiler. However, all
prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
been limited to simple, imperative languages.
